An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies

### 2017-05-17 main.tex
More updates

\begin{abstract}
    We present insights and empirical results from an extensive numerical
    study of the evolutionary dynamics of the iterated prisoner's dilemma.
    Fixation probabilities for Moran processes are obtained
    for all pairs of 164 different strategies including classics such as TitForTat, zero
    determinant strategies, and many more sophisticated strategies.
    Players with long memories and sophisticated behaviours outperform
    many strategies that perform well in a two player setting. Moreover we
    introduce several strategies trained with evolutionary algorithms to
    excel at the Moran process. These strategies are excellent invaders
    and resistors of invasion and in some cases naturally evolve handshaking
    mechanisms to resist invasion. The best invaders were those trained
    to maximize total payoff while the best resistors invoke handshake machanisms.
    This suggests that while maximizing individual payoff
    can lead to the evolution of cooperation through invasion, the relatively
    weak invasion resistance of payoff maximizing strategies are not as
    evolutionarily stable as strategies employing handshake mechanisms.
\end{abstract}

@@

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic}. The points represent the simulated
values and the line shows the theoretical value. Note that these are all
deterministic strategies and show a perfect match up between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategies pairs. Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of assuming a given interaction between two IPD strategies can be
summarised with a set of utilities as shown in
(\ref{equ:payoff_matrix}). For any given pair of strategies it is possible to
obtain \(p_{i,i-1}, p_{i,i+1}, p_{ii}\) exactly (as opposed to the
approximations offered by (\ref{equ:p_up}), (\ref{equ:p_down}) and
(\ref{equ:p_stay}). Obtaining these requires particular analysis for a given
pair and can be quite a complex endeavour for stochastic strategies with long
memory: this is not necessary for the purposes of this work.
All data generated for this validation exercise can be found
at \cite{data}.

@@

In this section the focus is on the ability of a mutant strategy to invade: the
probability of one individual of a given type successfully fixating in a
population of \(N - 1\) other individuals, denoted by \(x_1\).
The ranks of each strategy for all considered values of \(N\) according to mean
\(x_1\) are shown in Figure~\ref{fig:ranks_v_size_invade}.

@@

In addition to identifying good invaders, strategies resistant to invasion by
other strategies are identified by examining the distribution of $x_{N-1}$ for
each strategy. The ranks of each strategy for all considered values of \(N\)
according to mean \(x_{N-1}\) are shown in
Figures~\ref{fig:ranks_v_size_resist}.

@@

There are many opportunities build on this work. In particular, an analysis
of the effect of noise should offer insights regarding the stability of the findings,
particularly for the handshaking strategies. They may be less dominant for
larger amounts of noise since the handshaking mechanisms may become brittle.
There are many other variations to explore including populations with more
than one type, spatial structure, and mutation.

% TODO: author contributions

>>>>

\begin{abstract}
    In this paper an extensive numerical
    study of the evolutionary dynamics of the iterated prisoner's dilemma
    are presented. Fixation probabilities for Moran processes are obtained
    for 164 different strategies including classics such as TitForTat, zero
    determinant strategies, and many more sophisticated strategies.
    Players with long memories and sophisticated behaviours outperform
    many strategies that perform well in a two player setting. Moreover we
    introduce several strategies trained with evolutionary algorithms to
    excel at the Moran process. These strategies are excellent invaders
    and resistors of invasion and in some cases naturally evolve handshaking
    mechanisms to resist invasion.
\end{abstract}

@@

% With an intensity of selection \(\omega\) the fitness of both strategies is
% given by:
%
% \begin{equation}\label{equ:expected_payoff_one}
%     f_i = 1 - \omega + \omega F_i
% \end{equation}
%
% \begin{equation}\label{equ:expected_payoff_two}
%     g_i = 1 - \omega + \omega G_i
% \end{equation}

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic}. The points represent the simulated
values and the line shows the theoretical value. Note that these are all
deterministic strategies and show a perfect match up between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategies pairs. Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of the analytical formulae that relies on the average payoffs
(\ref{equ:payoff_matrix}). Estimating all the transitions may give better agreement
but is not necessary for our purposes. All data generated for this validation exercise can be found
at \cite{data}.

@@

In this section the focus is on the ability of a mutant strategy to invade: the
probability of one individual of a given type successfully fixating in a
population of \(N - 1\) other individuals, denoted by \(x_1\).
The ranks of each strategy for all considered values of \(N\) according to mean
\(x_1\) are shown in Figures~\ref{fig_ranks_v_size_invade}.

@@

In addition to identifying good invaders, strategies resistant to invasion by
other strategies are identified by examining the distribution of $x_{N-1}$ for
each strategy. The ranks of each strategy for all considered values of \(N\)
according to mean \(x_{N-1}\) are shown in
Figures~\ref{fig_ranks_v_size_resist}.

@@

% There are various areas for further work to build on this. Firstly, an analysis
% of the effect of noise would offer insights about the stability of the findings.
% It would also be possible to consider three or more types of strategy in the
% population and finally mutation would also offer an interesting dimension to
% explore.
